Collaborative Learning with Dirichlet Process Clustering for Rapid Online Adaptation in Robotics

### 2023-08-25 paper/paper.tex
Aug 25 version

\subsection{Gaussian Process}
The Gaussian Process (GP) is a powerful non-parametric machine learning algorithm. 
It assumes the prior distribution of the outcomes to be a multivariate normal distribution with a prior mean $\mu$ and a covariance matrix $\Sigma$. 
\begin{equation}
f(\bm{y}) = \frac{exp(
-\frac{1}{2}(\bm{y} - \bm{\mu})^T \bm{\Sigma^{-1}} (\bm{y} - \bm{\mu})
)}
{
\sqrt{(2\pi)^N det(\bm{\Sigma})}
}
\label{multi_normal}
\end{equation}
In contrast to regular multivariate normal distribution, GP allows the prior mean and the covariance matrix to be input dependent. The
prior mean becomes a function $\mu(\bm{x})$, and the covariance between two data points is given by a covariance function 
$k(\bm{x_1}, \bm{x_2})$ called kernel. 
Following the fact that data points closer to each other are more correlated, this covariance starts from the prior variance and asymptotically decreases to zero as the distance between the two points tends to infinity. 
To account for the fact the observations might be noisy, we can incorporate the noise into the covariance matrix by adding the variance of the noise on the diagonal.
\begin{equation}
\bm{\Sigma} = \bm{K}(\bm{x_{1:N}}, \bm{x_{1:N}}) + \sigma_n^2 \mathbf{I}
\label{covariance_matrix}
\end{equation}

This allows us to build reasonable joint prior distribution for the values of any well-behaved functions.
To make prediction with GP, we first use the prior mean function and kernel to build up the joint distribution as the prior. 
Then we can calculate the posterior for the data we wish to evaluate conditioned on the ones we have observed using Bayes rule.
Since the Gaussian distribution is a conjugate distribution, the posterior is still a Gaussian with mean and variance given by \cite{GP, GP_posterior}:
\begin{equation}
\mu_*(\bm{x}) = \mu(\bm{x}) + 
\bm{\Sigma_{N, x}}^T
\bm{\Sigma_{N, N}}^{-1}
(\bm{y_{1:N}} - \bm{\mu}(\bm{x_{1:N}}))
\label{posterior_mean}
\end{equation}
%
\begin{equation}
\sigma^2_*(\bm{x}) = \sigma^2(\bm{x}) -  
\bm{\Sigma_{N, x}}^T
\bm{\Sigma_{N, N}}^{-1}
\bm{\Sigma_{N, x}}
\label{posterior_var}
\end{equation}
where $\bm{\Sigma_{N, N}}$ denotes the covariance matrix of the observed data, and $\bm{\Sigma_{N, x}}$ denotes the column vector with entry on each row equal to the covariance between 
each observed data and the data we hope to evaluate.

@@

\subsection{Reset-free Trial-and-Error Learning}
The work of this paper is based on the Reset-free Trial-and-Error \cite{RTE} (RTE) algorithm. 
The RTE uses GP to learn the transformation model online.
This is a very clever choice for two reasons.
First, the online learning requires very high data-efficiency, which happens to be one of the strengths of GP. 
Secondly, the true dynamics happens at low level, while the repertoire-based method learns the transformation model at high level. This will inevitably introduce errors, which can be well incorporated by GP as being a probabilistic model.
In RTE, the task space is just the displacement of the robot in x and y direction after executing each policy.
The prior mean function of RTE is chosen to be the outcomes of the policies evaluated in the simulator, which are the results collected during the MAP-Elites policy generation. 
Since the policy outcome has two dimensions (x and y), RTE uses a different GP for each dimension.
The kernel is chosen to be the RBF kernel with an isotropic distance metric. The input space is chosen to be the same as the task space, namely the x and y displacement evaluated in the simulation. 
During online adaptation, the RTE first predicts the outcome of each policy in the repertoire using Eq. \ref{posterior_mean} and Eq. \ref{posterior_var}, then it uses a Monte Carlo Tree Search \cite{MCTS} (MCTS) to plan its actions.
It is worth noting that the RTE uses a periodic set of commands as the elementary policy. 
This means the policy execution is open-loop control, despite the robot knows its position and orientation.
While using such periodic controllers seems to be weaker than using a neural network, it is much more stable than using neural networks. 
This is because a periodic controller performs all its actions during evaluation, while neural networks are much more complicated and could behave very differently in different situations.
Since the repertoire based control aims to learn the distortion with limited number of interactions, the distortion cannot be too large or too complicated. Hence the use of periodic controller is a much better choice.

@@

p(\pi_1, \pi_2, \dots, \pi_k | \alpha) = 

@@

p(y_i|c_i = j, \bm{y}_{-i}, \bm{c}_{-i})

@@

p(c_1, c_2, \dots, c_n | \alpha) =  \frac{\Gamma(\alpha)}

@@

To calculate the last term in Eq. (\ref{indicator_posterior}), DP takes a full Bayesian approach:

@@

p(y_i|c_i = j, \bm{y}_{-i}, \bm{c}_{-i}) = 
\int p(y_i|\bm{\theta}_j)
p(\bm{\theta}_j|\bm{y}_{-i}, \bm{c}_{-i})
d\bm{\theta}_j

@@

p(\bm{\theta}_j|\bm{y}_{-i}, \bm{c}_{-i}) = \frac{1}{Z}
\,
p(\bm{\theta}_j)
\prod_{c_k=j} p(y_k|\bm{\theta}_j)

@@

where the $Z$ is a normalization factor. 

@@

The $-i$ subscript means we need to remove the data from its current cluster during resampling, which will affect the result of Eq. (\ref{integral_1}) when calculating the probability of the data remaining in its previous cluster. 

@@

\prod_{i=1}^n p(y_i|c_i = j, \bm{y}_i, \bm{c}_{-i})

@@

where the first product is made over each cluster containing at least one data. 
Note that the probability in the second product doesn't have the $-i$ subscript like in Eq. (\ref{integral_1}), so the data doesn't need to be removed from its current cluster now.

@@

\section{Methodology}
To formulate the problem clearly, we are aiming to leverage the collaboratively collected real-world interactions to assist a robot deployed in an unknown environment to quickly adapt and complete its task.
We are provided with some episodes (each task-solving process corresponds to an episode of interaction data) of real-world interactions collected by several robots deployed in an unknown distribution of environments.
The most intuitive idea is to label these historical data with their sources (like collected from grass terrain, collected with left leg broken, etc.) and build a simple database in the cloud server. 
During the adaptation, the robot examines the current environment and uploads the result to the database server. The server then identifies the most similar environment recorded in the database and let the robot download the corresponding data to assist the its adaptation.
However, accurately measuring and describing the environment can be expensive and difficult. It is also hard to determine whether a previously encountered environment shares similar dynamics with the current one even if we have accurate descriptions for them.
In the case we do know that some historical data come from exactly the same environment as the current one, these data might be insufficient in quantity to assist the current task.
Hence, it is reasonable to group the episodes resulting from similar dynamics into clusters, ensuring having enough data for analysis.
We will assume that we don’t have access to any information of the environment and aim to conduct clustering only according to the dynamics.

Previous works of clustering GP via DP have been made in \cite{infinite_MGP, variational_MGP}. 
But they work with individual data points, while in our case all the interaction data collected from the same task share the same indicator.
In this paper, we introduced a novel method to cluster batched data with infinite mixture of GPs while ensuring high flexibility.

There are a few assumptions needs to be made for further analysis. The first assumption is that the simulated results always correspond to real-world outcomes in some environment. 
For example, we set the friction coefficient to 0.9 in the simulation, but the it underestimated friction and the simulated outcomes are very close to the real-world results for friction coefficient equal to 0.8.
Hence we believe that although the simulated results are not accurate, they are still close to some real-world results.
This is not always true, but it plays a very important role in the methodology.

$$we \, have \, talked \, about \, prior \, mean \, importance$$
$$we \, have \, talked \, about \, simulation \, assumptions$$

@@

The prior mean function gives a prior estimate of the dynamics for a certain environment.
This means we need to estimate the outcomes of every elementary policy in this environment.
The dynamics can efficiently expressed in the following form:

@@

where the $x_d$ stands for the $d^{\text{th}}$ dimension of the policy rollout outcome (for example, this can be the robot displacement in the $x$ axis). Vector $\bm{\beta}$ represents the behaviour descriptor of the controller; vector $\bm{\epsilon}$ is the environment descriptor, representing some quantities of the environment. 

@@

A question is that how many distortions we should collect.
If we get too few distortions, we might not have collected all the basis, and the linear model will perform poorly in uncovered environments.
On the other hand, if we have collected too many, not only will this be computationally expensive but this will also lead to very large dimensions. 
The solution is collect as many as we can afford and then reduce the dimensionality with singular vector decomposition (SVD). The collection of distortions can be represented as:
\begin{equation}
\bm{A} = 
\begin{bmatrix}
\Delta \bm{x}_d^{(1)}, \cdots, \Delta \bm{x}_d^{(n)}
\end{bmatrix}
= \sum_{i=1}^s \sigma_i \bm{v}_i \bm{u}_i^T
\label{SVD}
\end{equation}
where $\bm{v}_i$ stands for each normalized base vector and $\bm{u}_i$ is the normalized component of this base vector in each distortion.
The number of basis $s$ is much smaller than $n$ as we have collected more than we need to ensure a good coverage.
To conduct SVD, we left multiply matrix $\bm{A}$ by its transpose:
\begin{equation}
\begin{gathered}
\bm{A}^T \bm{A} = 
\sum_{i=1}^s \sigma_i^2 \bm{u}_i \bm{u}_i^T = \bm{U} \bm{D} \bm{U}^T
\\
\text{where \,}
\bm{U} = 
\begin{bmatrix}
\bm{u}_1, \cdots, \bm{u}_s
\end{bmatrix}
\text{, \,}
\bm{D} = 
\begin{bmatrix}
\sigma_1^2 &  &  \\
 & \ddots &  \\
 &  & \sigma_s^2 \\
\end{bmatrix}
\end{gathered}
\label{ATA}
\end{equation}
According to spectral theorem, entry $\sigma^2$ and vectors $\bm{u}$  are the non-zero eigenvalues and the corresponding eigenvectors of $\bm{A}^T\bm{A}$, respectively. 
The desired base vectors can then be calculated:
\begin{equation}
\begin{gathered}
%\frac{1}{\sigma_j}\bm{A} \bm{u}_j = 
%\sum_{i=1}^s \frac{\sigma_i}{\sigma_j} \bm{v}_i \delta_{i,j}
%= \bm{v}_j
%\\
%\text{hence \,}
\begin{bmatrix}
\bm{v}_1, \cdots, \bm{v}_s
\end{bmatrix} = 
\sqrt{\bm{D}^{-1}} \bm{A} \bm{U}
\end{gathered}
\label{basis}
\end{equation}
The eigenvalues denote the importance of each base vector.
If an eigenvalue is zero, we can completely ignore this dimension as it can be linearly represented by other basis.
In practice, since evaluations are noisy and our linear relation is approximated, we will never get any zero eigenvalues. 
To continue with our dimensionality reduction, we can rearrange the eigenvalues in a descending order and keep only the first few terms depending on how many dimensions we wish to have.
Thus, we can approximate any real-world distortion using a linear combination of simulated distortions filtered by SVD.
Although this is just an approximation, it should serve well as the prior mean function of GP.

\subsection{Determine the Input Space}
The GP requires the input space for the kernel to build the covariance matrix.
In the RTE, this is the same as the $x$ and $y$ displacements.
Considering the policies are generated using the MAP-Elites instead of gradient based methods, policies having similar displacements might have very different behaviours.
As a result, they may lead to very different outcomes in new environments, and the use of such input space could be misleading.
Hence, we aim to find the input space in which the distances between policies are related to their outcomes in all environments.

@@

It is hard to determine the detailed linear combination to get the behaviour vectors.
But considering the use of anisotropic length scales for the distance metric in Eq. (\ref{distance_metric}), it is desirable to find a linear combination to reduce the correlations between the dimensions.
This is achieved using principle component analysis (PCA) \cite{PCA}. We first find the empirical covariance matrix of the standardized distortions.
\begin{equation}
\begin{gathered}
\bm{\Sigma} = \bm{X}^T \bm{X}
\end{gathered}
\label{cov}
\end{equation}
$\bm{X}$ is a $N \times s$ matrix, where $N$ is the size of the repertoire (the amount of elementary policies) and $s$ is the number of basis found using SVD.
Each column of $\bm{X}$, namely each dimension of basis, is standardized across the policies.
The covariance matrix $\bm{\Sigma}$ is a real-symmetric matrix, according to spectral theorem, it can be represented by:
\begin{equation}
\begin{gathered}
\bm{\Sigma} = 
\sum_{i=1}^s \lambda_i \bm{u}_i \bm{u}_i^T = \bm{U} \bm{D} \bm{U}^T
\end{gathered}
\label{PCA}
\end{equation}
Similar to SVD, the $\lambda_i$ are the eigenvalues of the covariance matrix, and $\bm{U}$ contains the corresponding eigenvectors.
%Similar to SVD, these eigenvalues are non-negative and give the importance of each dimension (see Fig. \ref{}).
We can then find our input space with the following transform:
\begin{equation}
\begin{gathered}
\begin{bmatrix}
\bm{v}'_1, \cdots, \bm{v}'_s
\end{bmatrix}
=
\begin{bmatrix}
\bm{v}_1, \cdots, \bm{v}_s
\end{bmatrix}
\bm{U} \sqrt{\bm{D}^{-1}}
\end{gathered}
\label{input_space}
\end{equation}
This gives the coordinates of each elementary policy in our input space. Each dimension of the coordinates are now decorrelated (see Fig. \ref{PCA_outcome}).
%
\begin{figure}[h]
\centering
\subfloat[]{\includegraphics[width=0.255\textwidth]{before_pca.pdf}%
\label{fig_first_case}}
%\hfil
\subfloat[]{\includegraphics[width=0.195\textwidth]{after_pca.pdf}%
\label{fig_second_case}}
\caption{Example of PCA transform. Original data distribution is given by (a), where the two axes are highly correlated. (b) is the resultant distribution after applying Eq. (\ref{input_space}), where the correlation is removed.}
\label{PCA_outcome}
\end{figure}

\subsection{Choosing the Dynamics Representation}
We can use different representations of the dynamics $x_d$ in Eq. (\ref{dynamics_assumption}).
Apart from using $x$ and $y$ displacements (called displacement-based representation in this paper) as in RTE, another type of representation is investigated.
Since RTE uses periodic controllers, the robot will experience the same displacement and rotation during each period, and the execution should result in an arc.
Based on this analysis, we represent the dynamics as the tangential speed $v$, angular velocity $\omega$ and direction $\varphi$ (see Fig. \ref{arcs}). 
This design, called arc-based representation, considers the fact that the different dimensions of the dynamics are modelled by different GPs (hence are decoupled) in RTE.
This is not an ideal design since $x$ and $y$ are typically correlated.
In arc-based representation, these parameters are more decoupled and more consistent with real-world mechanics. 
Take a hexapod robot for example.
If the robot is carrying heavy payload and becomes slower, arc-based representation can easily model this distortion by reducing $v$ for all policies. However, for displacement based representation, reduced speed will lead to decrease for positive displacements and increase for negative displacements.
In the case where the robot suffers damage in one of its legs, the side of the damaged leg will contribute lesser force, and the robot might move towards a different direction or rotate during the motion.
For arc-based representation, this can be modelled by adding a shift to the angular velocities and directions of the policies.
While it is a lot harder for the displacement-base representation, and $x$ and $y$ will become strongly correlated.
\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{example_curves.pdf}
\caption{Arc trajectories with different $v$, $\omega$ and $\varphi$. Curve 1 has $v=0.5$, $\omega = 30^\circ/s$ and $\varphi = 90^\circ$. Curve 2 has $v=0.25$, $\omega = 0$ and $\varphi = 90^\circ$.
Curve 3 shares the same $v$ and $\omega$ with curve 1 but having $\varphi = 0$.
All the curves are generated after 4 seconds of movement.
}
\label{arcs}
\end{figure}

@@

\subsection{Data preparation}

>>>>

\subsection{Gaussian Process}
The Gaussian Process (GP) is a powerful non-parametric machine learning algorithm. 
It assumes the prior distribution of the outcomes to be a multivariate normal distribution with a prior mean $\mu$ and a covariance matrix $\Sigma$. 
\begin{equation}
f(\bm{y}) = \frac{exp(
-\frac{1}{2}(\bm{y} - \bm{\mu})^T \bm{\Sigma^{-1}} (\bm{y} - \bm{\mu})
)}
{
\sqrt{(2\pi)^N \det(\bm{\Sigma})}
}
\label{multi_normal}
\end{equation}
In contrast to regular multivariate normal distribution, GP allows the prior mean and the covariance matrix to be input dependent. The
prior mean becomes a function $\mu(\bm{x})$, and the covariance between two data points is given by a covariance function 
$k(\bm{x_1}, \bm{x_2})$ called kernel. 
Following the fact that data points closer to each other are more correlated, this covariance starts from the prior variance and asymptotically decreases to zero as the distance between the two points tends to infinity. 
To account for the fact the observations might be noisy, we can incorporate the noise into the covariance matrix by adding the variance of the noise on the diagonal.
\begin{equation}
\bm{\Sigma} = \bm{K}(\bm{x_{1:N}}, \bm{x_{1:N}}) + \sigma_n^2 \mathbf{I}
\label{covariance_matrix}
\end{equation}

This allows us to build reasonable joint prior distribution for the values of any well-behaved functions.
To make prediction with GP, we first use the prior mean function and kernel to build up the joint distribution as the prior. 
Then we can calculate the posterior for the data we wish to evaluate conditioned on the ones we have observed using Bayes rule.
Since the Gaussian distribution is a conjugate distribution, the posterior is still a Gaussian with mean and variance given by \cite{GP, GP_posterior}:
\begin{equation}
\begin{gathered}
\mu_*(x) = \mu(x) + 
\bm{\Sigma}_{N, x}^T
\bm{\Sigma}_{N, N}^{-1}
(\bm{y}_{1:N} - \bm{\mu}(\bm{x}_{1:N}))
\\
\sigma^2_*(x) = \sigma^2(x) -  
\bm{\Sigma}_{N, x}^T
\bm{\Sigma}_{N, N}^{-1}
\bm{\Sigma}_{N, x}
\end{gathered}
\label{GP_posterior}
\end{equation}
where $\bm{\Sigma}_{N, N}$ denotes the covariance matrix of the observed data, and $\bm{\Sigma}_{N, x}$ denotes the column vector with entry on each row equal to the covariance between 
each observed data and the data we hope to evaluate.

@@

\subsection{Reset-free Trial-and-Error Learning}
The work of this paper is based on the Reset-free Trial-and-Error \cite{RTE} (RTE) algorithm. 
The RTE uses GP to learn the transformation model online.
This is a very clever choice for two reasons.
First, the online learning requires very high data-efficiency, which happens to be one of the strengths of GP. 
Secondly, the true dynamics happens at low level, while the repertoire-based method learns the transformation model at high level. This will inevitably introduce errors, which can be well incorporated by GP as being a probabilistic model.
In RTE, the task space is just the displacement of the robot in x and y direction after executing each policy.
The prior mean function of RTE is chosen to be the outcomes of the policies evaluated in the simulator, which are the results collected during the MAP-Elites policy generation. 
Since the policy outcome has two dimensions (x and y), RTE uses a different GP for each dimension.
The kernel is chosen to be the RBF kernel with an isotropic distance metric. The input space is chosen to be the same as the task space, namely the x and y displacement evaluated in the simulation. 
During online adaptation, the RTE first predicts the outcome of each policy in the repertoire using Eq. (\ref{GP_posterior}), then it uses a Monte Carlo Tree Search \cite{MCTS} (MCTS) to plan its actions.
It is worth noting that the RTE uses a periodic set of commands as the elementary policy. 
This means the policy execution is open-loop control, despite the robot knows its position and orientation.
While using such periodic controllers seems to be weaker than using a neural network, it is much more stable than using neural networks. 
This is because a periodic controller performs all its actions during evaluation, while neural networks are much more complicated and could behave very differently in different situations.
Since the repertoire based control aims to learn the distortion with limited number of interactions, the distortion cannot be too large or too complicated. Hence the use of periodic controller is a much better choice.

@@

p(\pi_1, \dots, \pi_k | \alpha) = 

@@

p(y_i|c_i=j, \bm{y}_{-i}, \bm{c}_{-i})

@@

p(c_1, \dots, c_n | \alpha) =  \frac{\Gamma(\alpha)}

@@

To calculate the last term in Eq. (\ref{indicator_posterior}), we integrate the likelihood over the posterior:

@@

p(y_i|c_i, \bm{y}_{-i}, \bm{c}_{-i}) = 
p(\bm{\theta}|c_i, \bm{y}_{-i}, \bm{c}_{-i})
d\bm{\theta}

@@

p(\bm{\theta}|c_i, \bm{y}_{-i}, \bm{c}_{-i}) \propto
\,\, p(\bm{\theta})
\prod_{c_k=c_i, k \neq i} p(y_k|\bm{\theta})

@@

@@

The $-i$ subscript on $\bm{y}$ means we need to remove the data from its current cluster during resampling, which will affect the result of Eq. (\ref{integral_1}) when calculating the probability of the data remaining in its current cluster. 

@@

\prod_{i=1}^n p(y_i|c_i, \bm{y}, \bm{c}_{-i})

@@

Note that the probability in the second product doesn't have the $-i$ subscript on $\bm{y}$ like in Eq. (\ref{integral_1}), so the data doesn't need to be removed from its current cluster now.

@@

\section{Methodology}
We are aiming to improve RTE by leveraging the collaboratively collected real-world interactions.
To be clear about the terminology, in repertoire-based method, an interaction means executing an certain elementary policy and acquiring its outcome.
DP enables us to achieve this by extending the GP model in RTE to an infinite mixture of GPs trained with real-world data.
Previous works of training infinite mixture of GP via DP have been made in \cite{infinite_MGP, variational_MGP}. 
But in our case, the design needs to be different.
First, our real-world interaction data are produced in individual episodes, where each episode corresponds to the interactions collected during each deployment.
Since we know that each deployment is made in a certain environment, all the interactions in an episode should correspond to the same dynamics. 
Hence, when clustering the historical data we need to conduct clustering on the episode but not on the individual data. 
The second difference from previous works is that we are decoupling each dimension and each episode.
The decoupling of dimension is inherited from RTE, while the decoupling of episodes is a unique design.
This is based on the fact that data within each episode come from the same environment and are hence correlated, while different episodes do not necessarily correspond to the same environment even they are grouped into the same cluster.
Such design allows us to group up environments that are different but similar, providing higher flexibility.
Another reason for using decoupled episodes is to make our system scalable.
Since we are now clustering episodes which typically contains dozens of data, each clusters can have thousands of data.
GP requires to compute the inverse of the joint covariance matrix, which has time complexity scaling cubic with data size.
Such computational expense is completely unaffordable for online adaptation. 
If the episodes are decoupled, the joint covariance matrix becomes a block diagonal matrix, and the time complexity of its inverse scales only linearly.

To really implement the DP clustering, there are many remaining challenges to be addressed.
The prior mean and kernel are the two most important factors of a GP.
It can be very hard to design a good prior function or to determine the length scales as we have little knowledge about the real-world dynamics.
A few parts of our methodology heavily rely on simulation.
For this, the simulation has to be relatively accurate and free of bugs.
We also make a very strong assumption that the simulated results always correspond to real-world outcomes in some environment. 
For example, we set the friction coefficient to 0.9 in the simulation, but the friction is underestimated and the simulated outcomes are very close to the real-world results for friction coefficient equal to 0.8.
Hence we believe that although the simulated results are not strictly accurate, they can still represent some real-world results.
This assumption not always true, but it plays a very important role in the methodology.

%To formulate the problem clearly, we are aiming to leverage the collaboratively collected real-world interactions to assist a robot deployed in an unknown environment to quickly adapt and complete its task.
%We are provided with some episodes (each task-solving process corresponds to an episode of interaction data) of real-world interactions collected by several robots deployed in an unknown distribution of environments.
%The most intuitive idea is to label these historical data with their sources (like collected from grass terrain, collected with left leg broken, etc.) and build a simple database in the cloud server. 
%During the adaptation, the robot examines the current environment and uploads the result to the database server. The server then identifies the most similar environment recorded in the database and let the robot download the corresponding data to assist the its adaptation.
%However, accurately measuring and describing the environment can be expensive and difficult. It is also hard to determine whether a previously encountered environment shares similar dynamics with the current one even if we have accurate descriptions for them.
%In the case we do know that some historical data come from exactly the same environment as the current one, these data might be insufficient in quantity to assist the current task.
%Hence, it is reasonable to group the episodes resulting from similar dynamics into clusters, ensuring having enough data for analysis.
%We will assume that we don’t have access to any information of the environment and aim to conduct clustering only according to the dynamics.

@@

In GP, the prior mean function gives a prior estimate of the outcomes of all the elementary policies in a certain environment.
To study the dynamics, we express its dependency in the following form:

@@

where the $x_d$ stands for the $d^{\text{th}}$ dimension of the policy outcome (for example, this can be the robot displacement in the $x$ axis). Vector $\bm{\beta}$ represents the behaviour descriptor of the controller; vector $\bm{\epsilon}$ is the environment descriptor, representing some quantities of the environment. 

@@

A question is that how many distortions we should collect.
If we get too few distortions, we might not have collected all the basis, and the linear model will perform poorly in uncovered environments.
On the other hand, if we have collected too many, not only will this be computationally expensive but this will also lead to very large dimensions. 
The solution is collect as many as we can afford and then reduce the dimensionality with singular vector decomposition (SVD). The collection of distortions can be represented as:
\begin{equation}
\bm{A} = 
\begin{bmatrix}
\Delta \bm{x}_d^{(1)}, \cdots, \Delta \bm{x}_d^{(n)}
\end{bmatrix}
= \sum_{i=1}^s \sigma_i \bm{v}_i \bm{u}_i^T
\label{SVD}
\end{equation}
where $\bm{v}_i$ stands for each normalized base vector and $\bm{u}_i$ is the normalized component of this base vector in each distortion.
The number of basis $s$ is much smaller than $n$ as we have collected more than we need to ensure a good coverage.
To conduct SVD, we left multiply matrix $\bm{A}$ by its transpose:
\begin{equation}
\begin{gathered}
\bm{A}^T \bm{A} = 
\sum_{i=1}^s \sigma_i^2 \bm{u}_i \bm{u}_i^T = \bm{U} \bm{D} \bm{U}^T
\\
\text{where \,}
\bm{U} = 
\begin{bmatrix}
\bm{u}_1, \cdots, \bm{u}_s
\end{bmatrix}
\text{, \,}
\bm{D} = 
\begin{bmatrix}
\sigma_1^2 &  &  \\
 & \ddots &  \\
 &  & \sigma_s^2 \\
\end{bmatrix}
\end{gathered}
\label{ATA}
\end{equation}
According to spectral theorem, entry $\sigma^2$ and vectors $\bm{u}$  are the non-zero eigenvalues and the corresponding eigenvectors of $\bm{A}^T\bm{A}$, respectively. 
The desired base vectors can then be calculated:
\begin{equation}
\begin{gathered}
%\frac{1}{\sigma_j}\bm{A} \bm{u}_j = 
%\sum_{i=1}^s \frac{\sigma_i}{\sigma_j} \bm{v}_i \delta_{i,j}
%= \bm{v}_j
%\\
%\text{hence \,}
\begin{bmatrix}
\bm{v}_1, \cdots, \bm{v}_s
\end{bmatrix} = 
\sqrt{\bm{D}^{-1}} \bm{A} \bm{U}
\end{gathered}
\label{basis}
\end{equation}
The eigenvalues denote the importance of each base vector.
If an eigenvalue is zero, we can completely ignore this dimension as it can be linearly represented by other basis.
In practice, since evaluations are noisy and our linear relation is approximated, we will never get any zero eigenvalues. 
To continue with our dimensionality reduction, we can rearrange the eigenvalues in a descending order and keep only the first few terms depending on how many dimensions we wish to have.
Thus, we can approximate any real-world distortion using a linear combination of simulated distortions filtered by SVD.

\subsection{Determine the Input Space}
In GP, the prior mean function only provides an estimate of the dynamics, and the residual is modelled with probabilistic inference.
This requires the definition of input space for the kernel to build the covariance between observations.
In the RTE, the input space the same as the $x$ and $y$ displacements.
Considering the policies are generated using the MAP-Elites instead of gradient based methods, policies having similar displacements might have very different behaviours.
As a result, they may lead to very different outcomes in new environments, and the use of such input space could be misleading.
Hence, we aim to find the input space in which the distances between policies are related to their outcomes in all environments.

@@

It is hard to determine the detailed linear combination to get the behaviour vectors.
But considering the use of anisotropic length scales for the distance metric in Eq. (\ref{distance_metric}), it is desirable to find a linear combination to reduce the correlations between the dimensions.
This is achieved using principle component analysis (PCA) \cite{PCA}. We first find the empirical covariance matrix of the standardized distortions.
\begin{equation}
\begin{gathered}
\bm{\Sigma} = \bm{X}^T \bm{X}
\end{gathered}
\label{cov}
\end{equation}
$\bm{X}$ is a $N \times s$ matrix, where $N$ is the size of the repertoire (the amount of elementary policies) and $s$ is the number of basis found using SVD.
Each column of $\bm{X}$, namely each dimension of basis, is standardized across the policies.
The covariance matrix $\bm{\Sigma}$ is a real-symmetric matrix, according to spectral theorem, it can be represented by:
\begin{equation}
\begin{gathered}
\bm{\Sigma} = 
\sum_{i=1}^s \lambda_i \bm{u}_i \bm{u}_i^T = \bm{U} \bm{D} \bm{U}^T
\end{gathered}
\label{PCA}
\end{equation}
Similar to SVD, the $\lambda_i$ are the eigenvalues of the covariance matrix, and $\bm{U}$ contains the corresponding eigenvectors.
%Similar to SVD, these eigenvalues are non-negative and give the importance of each dimension (see Fig. \ref{}).
We can then find our input space with the following transform:
\begin{equation}
\begin{gathered}
\begin{bmatrix}
\bm{v}'_1, \cdots, \bm{v}'_s
\end{bmatrix}
=
\begin{bmatrix}
\bm{v}_1, \cdots, \bm{v}_s
\end{bmatrix}
\bm{U} \sqrt{\bm{D}^{-1}}
\end{gathered}
\label{input_space}
\end{equation}
This gives the coordinates of each elementary policy in our input space. Each dimension of the coordinates are now decorrelated (see Fig. \ref{PCA_outcome}).
Since this coordinate system is just a linear transformation of the SVD basis with full rank, it can also be used as the basis for the prior mean function.
Hence, we replace the previous basis by this coordinate system for simplicity, and the distortion estimated by the prior mean function becomes: 
\begin{equation}
\begin{gathered}
\Delta \bm{x}_{d} = 
\bm{X}_{d}
\bm{w}_{\varepsilon, d}
\end{gathered}
\label{prior_mean_function}
\end{equation}
where each row of $\bm{X}_{d}$ is the coordinates (for the $d^{\text{th}}$ dimension of dynamics) of the corresponding policy.
%
\begin{figure}[h]
\centering
\subfloat[]{\includegraphics[width=0.255\textwidth]{before_pca.pdf}%
\label{fig_first_case}}
%\hfil
\subfloat[]{\includegraphics[width=0.195\textwidth]{after_pca.pdf}%
\label{fig_second_case}}
\caption{Example of PCA transform. Original data distribution is given by (a), where the two axes are highly correlated. (b) is the resultant distribution after applying Eq. (\ref{input_space}), where the correlation is removed.}
\label{PCA_outcome}
\end{figure}

\subsection{Choosing the Dynamics Representation}
We can use different representations of the dynamics $x_d$ in Eq. (\ref{dynamics_assumption}).
Apart from using $x$ and $y$ displacements (called displacement-based representation in this paper) as in RTE, another type of representation is investigated.
Since RTE uses periodic controllers, the robot will experience the same displacement and rotation during each period, and the execution should result in an arc.
Based on this analysis, we represent the dynamics as the tangential speed $v$, angular velocity $\omega$ and direction $\varphi$ (see Fig. \ref{arcs}). 
This design, called arc-based representation, considers the fact that the different dimensions of the dynamics are modelled by different GPs (hence are decoupled) in RTE.
This is not an ideal design since $x$ and $y$ are typically correlated.
In arc-based representation, these parameters are more decoupled and more consistent with real-world mechanics. 
Take a hexapod robot for example.
If the robot is carrying heavy payload and becomes slower, arc-based representation can easily model this distortion by reducing $v$ for all policies. However, for displacement-based representation, reduced speed will lead to decrease for positive displacements and increase for negative displacements.
In the case where the robot suffers damage in one of its legs, the side of the damaged leg will contribute lesser force, and the robot might move towards a different direction or rotate during the motion.
For arc-based representation, this can be modelled by adding a shift to the angular velocities and directions of the policies.
While it is a lot harder for the displacement-base representation, and $x$ and $y$ will become strongly correlated.
\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{example_curves.pdf}
\caption{Arc trajectories with different $v$, $\omega$ and $\varphi$. Curve 1 has $v=0.5$, $\omega = 30^\circ/s$ and $\varphi = 90^\circ$. Curve 2 has $v=0.25$, $\omega = 0$ and $\varphi = 90^\circ$.
Curve 3 shares the same $v$ and $\omega$ with curve 1 but having $\varphi = 0$.
All the curves are generated after 4 seconds of movement.
}
\label{arcs}
\end{figure}

@@

\subsection{Collecting Data Collaboratively}
